---
title: FE418 Proje
author: Kadir Yurdakul - 202301112
date: "`r Sys.Date()`"   
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    code_folding: show
    css: renk.css
---

# Functions

```{r}
check_fun <- function(parameter) {
  
  unique_length <- length(unique(parameter))
  
  parameter_length <- length(parameter)
  
  if (parameter_length == unique_length) {
    print(paste("Parameter length is equal to the number of unique values.", parameter_length, " = ", unique_length))
  } else {
    print(paste("Parameter length is not equal to the number of unique values.", parameter_length, " != ", unique_length))
  }
  
  if (all(!duplicated(parameter))) {
    print("The parameter does not contain duplicated values.")
  } else {
    print(paste("The parameter contains duplicated values.", length(duplicated(parameter)), "times"))
  }
}

```

```{r}
custom_summary <- function(data) {
  mean_val <- mean(data, na.rm = TRUE)
  median_val <- median(data, na.rm = TRUE)
  mode_val <- as.numeric(names(sort(-table(data)))[1])
  var_val <- var(data, na.rm = TRUE)
  sd_val <- sd(data, na.rm = TRUE)
  min_val <- min(data, na.rm = TRUE)
  max_val <- max(data, na.rm = TRUE)
  range_val <- max(data, na.rm = TRUE) - min(data, na.rm = TRUE)
  q1 <- quantile(data, na.rm = TRUE)[2]
  q3 <- quantile(data, na.rm = TRUE)[4]
  iqr_val <- q3 - q1
  skewness_val <- skewness(data, na.rm = TRUE)
  kurtosis_val <- kurtosis(data, na.rm = TRUE)
  
  summary_result <- list(
    "Mean" = mean_val,
    "Median" = median_val,
    "Mode" = mode_val,
    "Variance" = var_val,
    "Standard Deviation" = sd_val,
    "Minimum" = min_val,
    "Maximum" = max_val,
    "Range" = range_val,
    "1st Quartile" = q1,
    "3rd Quartile" = q3,
    "Interquartile Range" = iqr_val,
    "Skewness" = skewness_val,
    "Kurtosis" = kurtosis_val
  )
  
  return(summary_result)
}

```

```{r}
roman_to_int <- function(roman) {
  roman_numerals <- list(
    "0" = 0, 
    I = 1,
    V = 5,
    X = 10
  )
  
  total <- 0
  prev_value <- 0
  
  for (char in rev(strsplit(roman, "")[[1]])) {
    value <- roman_numerals[[char]]
    if (value >= prev_value) {
      total <- total + value
    } else {
      total <- total - value
    }
    prev_value <- value
  }
  
  return(total)
}

```

# Loading the Libraries

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(Amelia)
library(ggcorrplot)
library(tidyr)
library(ggtext)
library(e1071)

```

# Data examine n' import

```{r}

myfile <- list.files("data/", pattern = "\\.csv$", full.names = TRUE)
data_list <- lapply(myfile, read.csv);remove(myfile)
# We can access any data by data_list[[number]]
```

# Hypothesis 1: is sleep difficulty effect on sleep duration

## Data Check and Cleaning

```{r}
sleepdata <- data.frame(data_list[[1]]) # Assign the first set of data in data_list to a data frame called sleepdata.

```

```{r}
unique(sleepdata$sleep_difficulty)

sleepdata$sleep_difficulty <- gsub(paste("es", "yes", sep = "|"), "1", sleepdata$sleep_difficulty)

sleepdata$sleep_difficulty <- gsub(paste("no", "noo", "o", sep = "|"), "0", sleepdata$sleep_difficulty)

sleepdata$sleep_difficulty <- as.numeric(sleepdata$sleep_difficulty)

sleepdata$sleep_difficulty[sleepdata$sleep_difficulty == ""] <- NA

unique(sleepdata$sleep_difficulty)
```


```{r}
filtered_values <- sleepdata$sleep_duration[sleepdata$sleep_duration > 24 & !is.na(sleepdata$sleep_duration)]
sleepdata$sleep_duration[sleepdata$sleep_duration > 24 & !is.na(sleepdata$sleep_duration)] <- filtered_values * 0.1
remove(filtered_values)

unique(sleepdata$night_wakes)

sleepdata <- na.omit(sleepdata) 
head(sleepdata)
```


## Data Discription

```{r}
check_fun(sleepdata$ID)
```

```{r}
custom_summary(sleepdata$sleep_duration)
```

```{r}
qqnorm(sleepdata$sleep_duration)
qqline(sleepdata$sleep_duration)

```

```{r}
sum(sleepdata$sleep_difficulty[sleepdata$sleep_difficulty == 1])
sum(sleepdata$sleep_difficulty[sleepdata$sleep_difficulty == 0])

```

```{r,warning=FALSE}
ks.test(sleepdata$sleep_duration, "pnorm", mean = mean(sleepdata$sleep_duration), sd = sd(sleepdata$sleep_duration))

shapiro.test(sample(1:length(sleepdata$ID),4999))

```


```{r}
hist(sleepdata$sleep_duration, main = "Sleep Duration Histogram", xlab = "Duration", ylab = "Frequency", breaks = seq(1, 20, 1))

```

```{r}
barplot(table(sleepdata$sleep_duration),
        main = "Sleep Duration Frequency",
        xlab = "Sleep Duration",
        ylab = "Frequency",
        col = "lightblue",
        border = "blue")

```

```{r}
boxplot(sleepdata$sleep_duration)
```


## Data Comparison

```{r, warning=FALSE}
t_test_result <- t.test(sleepdata$sleep_duration ~ sleepdata$sleep_difficulty)
t_test_result;remove(t_test_result)

```

```{r}

corrplot::corrplot(cor(sleepdata[, sapply(sleepdata, is.numeric)], use = "complete.obs"),
                   method = "circle", type = "upper", order = "hclust", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45, 
         diag = FALSE)
```

```{r}
model <- lm(sleep_duration ~ night_wakes + sleep_difficulty, data = sleepdata)
  summary(model)

```

```{r}
sleepdata %>% ggplot(aes(sleep_duration, night_wakes, colour = sleep_difficulty)) +
  geom_point()+ 
  geom_smooth(method = "lm", se = FALSE, aes(group = sleep_difficulty)) 
```


```{r}
ggplot(sleepdata, aes(x = factor(sleep_difficulty), y = sleep_duration, colour = sleep_difficulty)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Sleep Duration by Sleep Difficulty", x = "Sleep Difficulty", y = "Sleep Duration") +
  scale_x_discrete(labels = c("0" = "Easy to Sleep", "1" = "Hard to Sleep"))

```

# Hypothesis 2: smoking effect on night wakes

## Data check n' cleaning

```{r}
socio <- data.frame(data_list[[2]])
socio <- left_join(socio, sleepdata)
socio <- socio %>% select(ID, smoke, sleep_difficulty)
```

```{r}
socio$smoke <- gsub("no", 0, socio$smoke)
socio$smoke <- gsub("past", 1, socio$smoke)
socio$smoke <- gsub("yes", 2, socio$smoke)
socio$smoke <- as.numeric(socio$smoke)
socio <- na.omit(socio)
head(socio)
```

## Data discription 

```{r}
check_fun(socio$ID)
```


```{r}
paste("num of smoking situation 0 is", sum(socio$smoke == 0, na.rm = T))
paste("num of smoking situation 1 is", sum(socio$smoke == 1, na.rm = T))
paste("num of smoking situation 2 is", sum(socio$smoke == 2, na.rm = T))
```

```{r}
paste("num of people has not sleep difficulty", sum(socio$sleep_difficulty == 0, na.rm = T))
paste("num of people has sleep difficulty", sum(socio$sleep_difficulty == 1, na.rm = T))

```

```{r}
smoke_percentage <- mean(socio$smoke > 0) * 100
cat("Sigara ienlerin yzdesi:", smoke_percentage, "%\n")

sleep_difficulty_percentage <- mean(socio$sleep_difficulty > 0) * 100
cat("Uyku gülüğü çekenlerin yzdesi:", sleep_difficulty_percentage, "%\n")

both_count <- sum(socio$ smoke > 0 & socio$sleep_difficulty > 0)
cat("Hem sigara iip hem de uyku gçl ekenlerin sayısı:", both_count, "\n")

```


```{r}
table(socio$smoke, socio$sleep_difficulty)
```



```{r,warning=FALSE}
ggplot(socio, aes(x = smoke)) +
  geom_bar() +
  labs(title = "Sleep Difficulty dist", x = "Sleep Difficulty", y = "Count") +
  theme_minimal()
```

```{r}
ggplot(socio, aes(x = smoke)) +
  geom_bar() +
  labs(title = "Smoke Situation dist", x = "Smoke Situation", y = "Count") +
  theme_minimal()
```

```{r}
mosaicplot(table(socio$sleep_difficulty, socio$smoke),
           main = "Sleep Difficulty ve Smoke Situation Arasındaki ",
           xlab = "Smoke Situation", ylab = "Sleep Difficulty", color = TRUE)

```

## data comparison

```{r}

anova_result <- aov(sleep_difficulty ~ smoke, data=socio)
anova_result
summary(anova_result)

```

```{r}
correlation_matrix <- cor(socio[, c("smoke", "sleep_difficulty")])

print(correlation_matrix)

```

```{r}
barplot(table(socio$smoke), main="Sigara", xlab="Sigara ", ylab="Frekans")

barplot(table(socio$sleep_difficulty), main="Uyku  kenlerin ", xlab="Uyku  ", ylab="Frekans")

```



```{r}
contingency_table <- table(socio$smoke, socio$sleep_difficulty)
chi_square_result <- chisq.test(contingency_table)
chi_square_result
```



# Hypothesis 3: smoking effect on mental and physical health .

```{r}
pathology <- data.frame(data_list[[3]])

pathology <- left_join(pathology, socio, by = "ID")
pathology <- pathology %>% select(ID,physical_health_score, smoke, mental_health_score)

```

## Data check n' cleaning


```{r}
non_smoke <- pathology %>% filter(smoke == 0)
past_smokee <- pathology %>% filter(smoke == 1)
now_smoke <- pathology %>% filter(smoke == 2)

pathology <- na.omit(pathology)
```


## data discription

```{r}
check_fun(pathology$ID)
```

```{r}
custom_summary(pathology$physical_health_score)
custom_summary(pathology$mental_health_score)
```


```{r}
qqnorm(pathology$mental_health_score)
qqline(pathology$mental_health_score)

qqnorm(pathology$physical_health_score)
qqline(pathology$physical_health_score)
```

```{r, warning=FALSE}
shapiro.test(pathology$physical_health_score)
shapiro.test(pathology$mental_health_score)

ks.test(pathology$physical_health_score, "pnorm", mean = mean(pathology$physical_health_score, na.rm = T), sd = sd(pathology$physical_health_score, na.rm = T))

ks.test(pathology$mental_health_score, "pnorm", mean = mean(pathology$mental_health_score, na.rm = T), sd = sd(pathology$mental_health_score, na.rm = T))

```

```{r}
hist(pathology$physical_health_score)
hist(pathology$mental_health_score)
```

```{r}
pathology %>% na.omit() %>% 
  ggplot(aes(x = factor(mental_health_score))) + 
  geom_bar() + 
  labs(title = "Mental Health Score Histogram", x = "Mental Health Score", y = "Count") +
  theme_minimal()

pathology %>% na.omit() %>% 
  ggplot(aes(x = factor(physical_health_score))) + 
  geom_bar() + 
  labs(title = "Mental Health Score Histogram", x = "Mental Health Score", y = "Count") +
  theme_minimal()
```

```{r}
boxplot(pathology$physical_health_score)
boxplot(pathology$mental_health_score)
```

## data compariosn

```{r}
anova_result1 <- aov(pathology$mental_health_score ~ factor(smoke, levels = c(0, 1, 2)), data=pathology)
anova_result
summary(anova_result)
```


```{r}
anova_result2 <- aov(pathology$physical_health_score ~ factor(smoke, levels = c(0, 1, 2)), data=pathology)
anova_result
summary(anova_result)
```

```{r}
group0 <- pathology %>% filter(smoke == 0)
group1 <- pathology %>% filter(smoke == 1)
group2 <- pathology %>% filter(smoke == 2)

t.test(group0$physical_health_score, group1$physical_health_score)
t.test(group1$physical_health_score, group2$physical_health_score)
t.test(group0$physical_health_score, group2$physical_health_score)

t.test(group0$mental_health_score, group1$mental_health_score)
t.test(group1$mental_health_score, group2$mental_health_score)
t.test(group0$mental_health_score, group2$mental_health_score)

```

```{r}
pathology %>% select(physical_health_score,smoke,mental_health_score) %>% cor()
  
  
corrplot::corrplot( pathology %>% select(physical_health_score,smoke,mental_health_score) %>% cor(),
                    method = "circle")
```


```{r}
chi_square_result2 <- chisq.test(pathology$smoke, pathology$mental_health_score)
chi_square_result1 <- chisq.test(pathology$smoke, pathology$physical_health_score)

print(chi_square_result1)
print(chi_square_result2)

```

```{r}
ggplot() +
  geom_boxplot(data = group0, aes(x = "Group 0", y = physical_health_score), fill = "lightblue") +
  geom_boxplot(data = group1, aes(x = "Group 1", y = physical_health_score), fill = "lightgreen") +
  geom_boxplot(data = group2, aes(x = "Group 2", y = physical_health_score), fill = "lightpink") +
  labs(x = "Smoke Group", y = "Physical Health Score", title = "Physical Health Score by Smoke Group")
```


```{r}
ggplot() +
  geom_boxplot(data = group0, aes(x = "Group 0", y = mental_health_score), fill = "lightblue") +
  geom_boxplot(data = group1, aes(x = "Group 1", y = mental_health_score), fill = "lightgreen") +
  geom_boxplot(data = group2, aes(x = "Group 2", y = mental_health_score), fill = "lightpink") +
  labs(x = "Smoke Group", y = "Mental Health Score", title = "Mental Health Score by Smoke Group")
```

```{r}
model <- lm(mental_health_score ~ physical_health_score, data = pathology)

summary(model)
```

```{r}
# Scatterplot ve regresyon doğrusunu oluştur
ggplot(pathology, aes(x = physical_health_score, y = mental_health_score)) +
  geom_point() +  # Scatterplot
  geom_smooth(method = "lm", se = FALSE) +  # Regresyon doğrusu
  labs(title = "Regression Analysis: Mental Health vs. Physical Health",
       x = "Physical Health Score",
       y = "Mental Health Score") +
  theme_minimal()

```



# Hypothesis 4: sleep duraiton effect on body mass index

## Data check and cleaning

```{r}
bmi <- data_list[[4]]
bmi <- left_join(bmi, sleepdata)
bmi <- bmi %>% select(ID,bmi,sleep_duration)
```

```{r}
bmi_high <- bmi %>% filter(sleep_duration >= mean(sleep_duration, na.rm = T))
bmi_low <- bmi %>% filter(sleep_duration < mean(sleep_duration, na.rm = T))
```


## Data Discription

```{r}
check_fun(bmi$ID)
```

```{r}
custom_summary(bmi$bmi)
```

```{r}
qqnorm(bmi$bmi)
qqline(bmi$bmi)

```

```{r}
shapiro.test(bmi$bmi)

ks.test(bmi$bmi, "pnorm", mean = mean(bmi$bmi, na.rm = T), sd= sd(bmi$bmi,na.rm = T))
```

```{r}
hist(bmi$bmi)
```

```{r}
barplot(table(bmi$bmi),
        main = "BMI Frequency",
        xlab = "BMI",
        ylab = "Frequency",
        col = "lightblue",
        border = "blue")

```

```{r}
boxplot(bmi$bmi)
```

## Data comparison

```{r}
t.test(bmi_high$sleep_duration, bmi_low$sleep_duration)

```
```{r}
bmi %>% na.omit() %>% select(bmi,sleep_duration) %>% cor()
```

```{r}
bmi_for_cor <- bmi %>%
  select(-ID) %>%
  mutate(bmi_high = ifelse(bmi > mean(bmi, na.rm = TRUE), 1, 0)) %>%
  mutate(bmi_low = ifelse(bmi <= mean(bmi, na.rm = TRUE), 1, 0))
remove(bmi_high, bmi_low)
correlation_matrix <- cor(bmi_for_cor, use = "complete.obs")
ggcorrplot(correlation_matrix, lab = TRUE)
```

```{r}

bmi_for_cor <- bmi_for_cor %>%
  mutate(category = ifelse(bmi_high == 1, "High", "Low"))

ggplot(data = bmi_for_cor, aes(x = category, y = bmi, fill = category)) +
  geom_boxplot() +
  labs(x = "Category", y = "BMI", title = "Boxplot of BMI for High and Low Categories") +
  scale_fill_manual(values = c("red", "blue"))  # Renkleri belirleme
remove(bmi_for_cor, correlation_matrix)
```

```{r}
bmi %>% 
  ggplot(aes(bmi)) +
  geom_density() +
  labs(x = "BMI", y = "Sleep Duration", title = "Density Plot of BMI and Sleep Duration")

bmi %>% 
  ggplot(aes(sleep_duration)) +
  geom_density() +
  labs(x = "BMI", y = "Sleep Duration", title = "Density Plot of BMI and Sleep Duration")
```

# Hypothesis 5: sleep duraiton effect on cortisol level

## data check cleaning

```{r}
cortisol <-data.frame(data_list[[5]])
cortisol <- left_join(cortisol, sleepdata)
cortisol <- cortisol %>% select(ID,cortisol_9am,cortisol_10am,cortisol_11am,cortisol_14pm,cortisol_18pm,cortisol_22pm,sleep_duration)
```

```{r}
cortisol <- na.omit(cortisol)

cortisol_long <- cortisol %>%
  pivot_longer(cols = starts_with("cortisol_"), 
               names_to = "time", 
               values_to = "cortisol_level") %>%
  mutate(time = case_when(
    time == "cortisol_9am" ~ "09",
    time == "cortisol_10am" ~ "10",
    time == "cortisol_11am" ~ "11",
    time == "cortisol_14pm" ~ "14",
    time == "cortisol_18pm" ~ "18",
    time == "cortisol_22pm" ~ "22"
  ))
cortisol_long$sleep_situation <- ifelse(cortisol_long$sleep_duration >= mean(cortisol_long$sleep_duration, na.rm = TRUE), "more", "less")

cortisol$sleep_situation <- ifelse(cortisol$sleep_duration >= mean(cortisol$sleep_duration, na.rm = TRUE), "more", "low")

cortisol_more <- cortisol %>% filter(sleep_situation == "more")
cortisol_low <- cortisol %>% filter(sleep_situation == "low")

```

## data discription

```{r}
check_fun(cortisol$ID)
```

```{r}
custom_summary(cortisol$cortisol_9am)
custom_summary(cortisol$cortisol_10am)
custom_summary(cortisol$cortisol_11am)
custom_summary(cortisol$cortisol_14pm)
custom_summary(cortisol$cortisol_18pm)
custom_summary(cortisol$cortisol_22pm)
```

```{r}
hist(cortisol_long$cortisol_level, main = "Kortizol Seviyesi Dağılımı")
boxplot(cortisol_long$cortisol_level, main = "Kortizol Seviyesi Kutu Grafiği")

```

```{r}
shapiro.test(cortisol$cortisol_9am)$p.value
shapiro.test(cortisol$cortisol_10am)$p.value
shapiro.test(cortisol$cortisol_11am)$p.value
shapiro.test(cortisol$cortisol_14pm)$p.value
shapiro.test(cortisol$cortisol_18pm)$p.value
shapiro.test(cortisol$cortisol_22pm)$p.value
```


## data comparison


```{r}

ggplot(cortisol_long, aes(x = time, y = cortisol_level, fill = sleep_situation)) +
  geom_boxplot() +
  labs(x = "Saat", y = "Kortizol Seviyesi", fill = "Uyku Durumu") +
  scale_fill_manual(values = c("more" = "blue", "less" = "red")) +  
  facet_wrap(~time, scales = "free") +
  theme_minimal()

```


```{r}
model <- aov(cortisol_level ~ time + sleep_situation, data =cortisol_long )
model
```

```{r}
ggplot(data = cortisol_long, aes(x = time, y = cortisol_level, color = sleep_situation)) +
  geom_point() +
  labs(x = "Zaman", y = "Kortizol Seviyesi", color = "Uyku Durumu") +
  ggtitle("Zamana Gore Kortizol Seviyeleri ve Uyku Durumu")
```


```{r}
ggplot(data = cortisol_long, aes(x = time, y = cortisol_level, color = sleep_situation)) +
  geom_point() +
  labs(x = "Zaman", y = "Kortizol Seviyesi", color = "Uyku Durumu") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, aes(group = sleep_situation)) +
  ggtitle("Zamana Gore Kortizol Seviyeleri ve Uyku Durumu")
```





# Hypothesis 6: sleep duration effect on cognitive

## data check and clean

```{r}
cognitive <- data.frame(data_list[[6]])
```

```{r}
cognitive <- left_join(cognitive, sleepdata, by = "ID")

cognitive <- cognitive %>% select(ID, cognitive_error_1,cognitive_error_2, sleep_duration)

# cognitive1 <- cognitive %>%
#   mutate(sleep_situation = ifelse(sleep_duration >= mean(sleep_duration, na.rm = TRUE), 1, 2))

cognitive <- cognitive %>%
  mutate(sleep_situation = ifelse(sleep_duration >= mean(sleep_duration, na.rm = TRUE), 1, 0))

cognitive <- cognitive %>%
  mutate(mean_error = rowMeans(select(., cognitive_error_1, cognitive_error_2), na.rm = TRUE))

cognitive <- cognitive %>% select(ID, mean_error,,sleep_situation,sleep_duration)

```

```{r}
cognitive_low <- cognitive %>% filter(sleep_situation == 0)
cognitive_high <- cognitive %>% filter(sleep_situation == 1)
```

## data discription

```{r}
check_fun(cognitive$ID)
```


```{r}
custom_summary(cognitive$mean_error)

paste("num of high sleep level", sum(cognitive_high$sleep_duration, na.rm = T))
paste("num of low sleep level", sum(cognitive_low$sleep_duration, na.rm = T))
```

```{r}
qqnorm(cognitive$mean_error)
qqline(cognitive$mean_error)
```

```{r}
hist(cognitive$mean_error)

```

```{r}
shapiro.test(cognitive$mean_error)
```

## data comparison

```{r}
t.test(cognitive_high$mean_error, cognitive_low$mean_error)

```

```{r}
correlation_matrix <- cor(cognitive %>% select(cognitive_error_1, cognitive_error_2, sleep_duration),
                          use = "complete.obs")
correlation_matrix
```

```{r}
ggcorrplot(correlation_matrix, 
           lab = TRUE, 
           title = "Korelasyon Matrisi")
```


```{r}
model <- lm(mean_error ~ sleep_duration, data = cognitive)
summary(model)
```

```{r}
ggplot(cognitive, aes(x = sleep_duration, y = mean_error)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Uyku Suresi", y = "Bilissel Hata 1", title = "Uyku Suresi ve Bilissel Hata 1 Arasındaki Lineer Regresyon") +
  theme_minimal()
```
```{r}
cognitive %>% na.omit() %>% ggplot(aes(x = as.factor(sleep_duration), y = mean_error, fill = sleep_situation)) +
  geom_boxplot() +
  labs(title = "Sleep Duration vs Mean Error by Sleep Situation",
       x = "Sleep Duration",
       y = "Mean Error",
       fill = "Sleep Situation") +
  theme_minimal()
```



# Hypothesis 7: sleep duration vs pathology 

## data clean and check

```{r}
pathology <- data.frame(data_list[[7]])
pathology$Alzheimer_score[pathology$Alzheimer_score == ""] <- NA
pathology <- na.omit(pathology)

```

```{r}
unique(pathology$Alzheimer_score)
```


```{r}
pathology$Alzheimer_score_num <- NA

```

```{r}
for (i in 1:length(pathology$Alzheimer_score)) {
  
  parts <- strsplit(pathology$Alzheimer_score[i], "-")[[1]]
  len <- length(parts)
  
  if (len == 1) {
    pathology$Alzheimer_score_num[i] <- roman_to_int(parts[1])
  } else {
    value1 <- roman_to_int(parts[1])
    value2 <- roman_to_int(parts[2])
    pathology$Alzheimer_score_num[i] <- (value1 + value2) / 2
  }
}

remove(i,len,parts, value1, value2)
```

```{r}
pathology <- left_join(pathology, sleepdata)
pathology <- pathology %>% select(ID,Alzheimer_score_num,sleep_duration)
```


```{r}

pathology <- pathology %>%
  mutate(sleep_situation = ifelse(sleep_duration >= mean(sleep_duration, na.rm = TRUE), 1, 0))

pathology_low <- pathology %>% filter(sleep_situation == 0)
pathology_high <- pathology %>% filter(sleep_situation == 1)
pathology <- na.omit(pathology)
```

## data discription

```{r}
str(pathology)
```

```{r}
custom_summary(pathology$Alzheimer_score_num)
custom_summary(pathology$sleep_duration)
```

```{r}
check_fun(pathology$ID)
```
```{r}
hist(pathology$Alzheimer_score_num)

```

```{r}
shapiro.test(pathology$Alzheimer_score_num)
```

## data comparison

```{r}
korelasyon_matrix <- cor(pathology$Alzheimer_score_num, pathology$sleep_duration, use = "complete.obs")
print(korelasyon_matrix)

```


```{r}
t_test_res <- t.test(pathology_high$Alzheimer_score_num, pathology_low$Alzheimer_score_num)
t_test_res
```


```{r}
ggplot() +
  geom_boxplot(data = pathology_low, aes(x = sleep_situation, y = Alzheimer_score_num, fill = sleep_situation), width = 0.5) +
  geom_boxplot(data = pathology_high, aes(x = sleep_situation, y = Alzheimer_score_num, fill = sleep_situation), width = 0.5) +
  labs(x = "Sleep Situation", y = "Alzheimer_score_num", title = "Comparison of Sleep Duration by Sleep Situation") +
  scale_fill_manual(values = c("less" = "blue", "more" = "red")) +
  theme_minimal()
```

