---
title:  Research Project
author: Kadir Yurdakul - 280104069
date: "`r Sys.Date()`"   
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    code_folding: show
    theme: readable
---

```{r message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(ggcorrplot)
library(corrplot)
library(hrbrthemes)
library(viridis)

```

# Data import and preperation for global data

```{r}
Socio_data <- read.csv(file = "C:/Users/Kadir/Desktop/Research Project1/raw/raw/PersonalDetailsQuestionnaire_04-2021.csv")
depression_data <- read.csv(file = "C:/Users/Kadir/Desktop/Research Project1/raw/raw/DepressionData_04-2021.csv")
stq_Data <- read.csv(file = "C:/Users/Kadir/Desktop/Research Project1/raw/raw/SleepTimingQuestionnaire_04-2021.csv")
health_data <- read.csv(file = "C:/Users/Kadir/Desktop/Research Project1/raw/raw/CornellMedicalIndex_04-2021.csv")
slp_data <- read.csv(file = "C:/Users/Kadir/Desktop/Research Project1/raw/raw/PersonalDetailsQuestionnaire_SleepEat_04-2021.csv")

```

## Column select for each datasets

```{r}
Socio_data <- Socio_data %>% 
  select(ID, sex, soclass, isced97total, p1mstat, p1curocc, p1cuhlth, p1illnes, p1smoke, p1didsmk, p1yrsmk, p1dayhos, p1drink, p1evdrnk, p1yrdrnk, p1dlacta, p1dlactb, p1dlactc, p1dlactd, p1dlacte, p1dlactf,p1dlactg,p1dlacth,p1dlacti, p1frnds1)
```

```{r}
depression_data <- depression_data %>% 
  select(ID, m1date, m1age,m1score,mood1,mood1sev)
```

```{r}
stq_Data <- stq_Data %>% 
  select(ID, stq_age, ,weight_stq,bmi_stq,socialjetlag,msfsc_stq,msf_stq)
```

```{r}
health_data <- health_data %>% 
  select(ID, cmiage1,cmi1_tot1,cmi1_tot2)
```

```{r}
slp_data <- slp_data %>% 
  select(ID,p1age,p1getup,p1slpnt,p1hrslp)
```

## Combine all data as full_data

```{r}
full_data <- full_join(Socio_data, depression_data ,by = "ID")
full_data <- full_join(full_data,slp_data ,by = "ID")
full_data <- full_join(full_data,health_data ,by = "ID")
full_data <- full_join(full_data, stq_Data ,by = "ID")

full_data <- full_data %>% 
  select(ID, sex, m1age, p1age,stq_age,cmiage1, everything())

full_data %>% na.omit() %>% head()

```

# Data Type and Cleaning

```{r}
str(full_data)
```

## Rearrange for age column

```{r}
full_data <- full_data %>% 
  mutate(age_avg = round((m1age + p1age) / 2, 0)) %>% 
  select(ID, sex, age_avg, everything()) %>% 
  select(-m1age, -p1age)

head(full_data$age_avg)
```

## Rearrange for sex column

```{r}
unique(full_data$sex)


full_data <- full_data %>%
  mutate(sex = na_if(sex, ""))

full_data$sex <- factor(full_data$sex, levels = c("Female", "Male"))

unique(full_data$sex)

```

## Rearrange for isced97total column

```{r}
unique(full_data$isced97total)

level_for_isced <- c("tertiary education first stage level 5", 
          "post secondary non-tertiary education level 4",
          "upper secondary education level 3",
          "lower secondary education level 2", 
          "tertiary education second stage")

full_data$isced97total <- factor(full_data$isced97total, levels = level_for_isced)
head(
  full_data$isced97total
)
```

## Rearrange for socialclass column

```{r}
unique(full_data$soclass)

full_data$soclass[full_data$soclass == ""] <- NA 


occupation_levels <- c(
  "I - Professional occupations",
  "II - Intermediate occupations",
  "III(M) - Skilled Manual occupations",
  "III(N) - Skilled Non-manual occupations",
  "IV - Partly Skilled occupations",
  "V - Unskilled occupations"
)


full_data$soclass <- factor(full_data$soclass, levels = occupation_levels, ordered = TRUE)


full_data <- full_data %>% 
  select(ID,sex,age_avg,soclass,,everything())

unique(full_data$soclass)

```

## Rearrange for marriage status column

```{r}
unique(full_data$p1mstat)


full_data$p1mstat[full_data$p1mstat == ""] <- NA
full_data$p1mstat[full_data$p1mstat == "widowed"] <- "divorced"
full_data$p1mstat[full_data$p1mstat == "separated"] <- "divorced"


full_data <- full_data %>% 
  mutate(
    p1mstat = factor(
      p1mstat, 
      levels = c("Married", "divorced", "single")
    ))

unique(full_data$p1mstat)

```

## Rearrange for current occupation column

```{r}
unique(full_data$p1curocc)

full_data$p1curocc[full_data$p1curocc == ""] <- NA

full_data <- full_data <- full_data %>% 
  mutate(
    p1curocc = case_when(
      p1curocc == "retired" | p1curocc == "housewife" | p1curocc == "unemployed" ~ "not-working",
      p1curocc == "part-time employment" | p1curocc == "retired with part-tme employment" | p1curocc == "housewife with a part-time job" ~ "part-time-working",
      p1curocc == "retired with a full time job" | p1curocc == "full-time employment" ~ "full-time-working",
      TRUE ~ p1curocc  # This ensures other values are retained
    )
  ) 



employment_levels <- c(
  "full-time-working",
  "part-time-working",
  "not-working"
)

full_data$p1curocc <- factor(full_data$p1curocc, levels = employment_levels, ordered = TRUE)


unique(full_data$p1curocc)

```

## Rearrange for current health column

```{r}
unique(full_data$p1cuhlth)
full_data$p1cuhlth <- gsub("Very bad", 0, full_data$p1cuhlth)
full_data$p1cuhlth <- gsub("Bad", 1, full_data$p1cuhlth)
full_data$p1cuhlth <- gsub("Fair", 2, full_data$p1cuhlth)
full_data$p1cuhlth <- gsub("Good", 3, full_data$p1cuhlth)
full_data$p1cuhlth <- gsub("Very good", 4, full_data$p1cuhlth)

full_data$p1cuhlth[full_data$p1cuhlth == ""] <- NA
full_data$p1cuhlth <- as.numeric(full_data$p1cuhlth)

unique(full_data$p1cuhlth)

```

## Rearrange for illnes column

```{r}
unique(full_data$p1illnes)

full_data$p1illnes <- gsub("yes", 1, full_data$p1illnes)
full_data$p1illnes <- gsub("no", 0, full_data$p1illnes)

full_data$p1illnes[full_data$p1illnes == ""] <- NA
full_data$p1illnes <- as.numeric(full_data$p1illnes)

unique(full_data$p1illnes)

```

## Rearrange and combine for smoke columns

```{r}
unique(full_data$p1smoke)
unique(full_data$p1didsmk)

full_data$p1smoke[full_data$p1smoke == ""] <- NA
full_data$p1didsmk[full_data$p1didsmk == ""] <- NA

unique(full_data$p1smoke)
unique(full_data$p1didsmk)


```

```{r}
full_data <- full_data %>%
  mutate(
    smoke = case_when(
      p1smoke == "No" & is.na(p1didsmk) & p1yrsmk == 0.00 ~ 0,  # never
      p1smoke == "No" & is.na(p1didsmk) & p1yrsmk > 0.00 ~ 1,   # smoke before
      p1smoke == "No" & is.na(p1didsmk) & p1yrsmk == is.na(p1yrsmk) ~ 0,  # never
      
      p1smoke == "No" & p1didsmk == "No" ~ 0,                  # never
      p1smoke == "No" & p1didsmk == "Yes" ~ 1,                 # smoek before
      p1smoke == "Yes" ~ 2                                     # still smoke
    )
  )

full_data %>% 
  select(p1smoke, p1didsmk, smoke, p1yrsmk, age_avg) %>% 
  filter(is.na(p1smoke)) %>% 
  head()

full_data %>% 
  select(p1smoke, p1didsmk, smoke, p1yrsmk, age_avg) %>% 
  head()

```

## Rearrange and combine for drink column

```{r}
full_data %>% select(p1drink,p1evdrnk,p1yrdrnk, age_avg) %>% 
  head()

unique(full_data$p1drink);unique(full_data$p1evdrnk)



full_data$p1drink[full_data$p1drink == ""] <- NA
full_data$p1evdrnk[full_data$p1evdrnk == ""] <- NA

full_data$p1evdrnk[full_data$p1evdrnk == "no"] <- "No"
full_data$p1evdrnk[full_data$p1evdrnk == "yes"] <- "Yes"


full_data <- full_data %>%
  mutate(
    drink = case_when(
      p1drink == "No" & is.na(p1evdrnk) & is.na(p1yrdrnk) ~ 0,  # Never
      p1drink == "No" & is.na(p1evdrnk) & p1yrdrnk == 0.00 ~ 0, # Never
      p1drink == "No" & is.na(p1evdrnk) & p1yrdrnk > 0.00 ~ 1,  # Drink before
      p1drink == "No" & p1evdrnk == "No" ~ 0,                  # Never
      p1drink == "No" & p1evdrnk == "Yes" ~ 1,                 # Drink before
      p1drink == "Yes" ~ 2                                     # Still drink
    )
  )

full_data %>% select(p1drink,p1evdrnk,p1yrdrnk, age_avg, drink) %>% head()

```

## Rearrange and make dlact score for dlact columns

```{r}
unique(full_data$p1dlacta)

full_data <- full_data %>%
  mutate(across(
    starts_with("p1dlact"), 
    ~ case_when(
      . == "" ~ NA,  
      . == "no difficulty" ~ 0,
      . == "some difficulty" ~ 1,
      . == "great difficulty" ~ 2,
    )
  )) %>%
  mutate(
    dlact_score = p1dlacta + p1dlactb + p1dlactc + p1dlactd + p1dlacte + p1dlactf + p1dlactg + p1dlacth + p1dlacti
  )


full_data %>% select(p1dlacta,p1dlactb,p1dlactc,p1dlactd,p1dlacte,p1dlactf,p1dlactg,p1dlacth,p1dlacti, dlact_score) %>% head()
unique(full_data$p1dlacta)

```

## Rearrange for frineds column

```{r}
unique(full_data$p1frnds1)

full_data$p1frnds1[full_data$p1frnds1 == ""] <- NA

full_data <- full_data %>% 
  mutate(
    p1frnds1 = case_when(
      p1frnds1 == "Less than average" ~ 0,
      p1frnds1 == "Average" ~ 1,
      p1frnds1 == "More than average" ~ 2
      
    )
  )
unique(full_data$p1frnds1)

```

## Rearrange for mood columns

```{r}
unique(full_data$mood1)
unique(full_data$mood1sev)

full_data$mood1[full_data$mood1 == ""] <- NA
full_data$mood1sev[full_data$mood1sev == ""] <- NA

full_data <- full_data %>% 
  mutate(
    mood1 = case_when(
      mood1 == "No" ~ 0,
      mood1 == "Depression" ~ 1
    )
  )

full_data <- full_data %>% 
  mutate(
    mood1sev = case_when(
      mood1sev == "No" ~ 0,
      mood1 == "Depression" ~ 1
    )
  )


```

```{r}
full_data <- full_data %>% select(-p1dlacta,-p1dlactb,-p1dlactc,-p1dlactd,-p1dlacte,-p1dlactf,-p1dlactg,-p1dlacth,-p1dlacti)

full_data <- full_data %>% select(ID,sex,age_avg,soclass,isced97total,dlact_score,drink,smoke,mood1,mood1sev,p1illnes,p1cuhlth,p1mstat, everything())
```

# Data Discription

## full_data

```{r}
head(full_data)
```

```{r}
str(full_data)
```

```{r}
dim(full_data)
```

```{r}
glimpse(full_data)

```

## corplot for fulldata
```{r}
numeric_data <- full_data %>% select_if(is.numeric)

correlation_matrix <- cor(numeric_data, method = "spearman", use = "complete.obs")

corrplot(correlation_matrix, method = "circle")

```

```{r}
ggcorrplot(correlation_matrix, lab = F)

```

```{r}
correlation_matrix
```

## how many people smoke and drink

```{r}

smoking <- full_data %>% filter(smoke >= 1) %>% summarise(count = n())
non_smoking <- full_data %>% filter(smoke == 0) %>% summarise(count = n())

smoking;non_smoking

drinking <- full_data %>% filter(drink >= 1) %>% summarise(count = n())
non_drinking <- full_data %>% filter(drink < 1) %>% summarise(count = n())

drinking;non_drinking

smoke_and_drink <- full_data %>% filter(smoke >= 1, drink >= 1) %>% summarise(count = n())
(smoke_and_drink$count / smoking$count) * 100 # What percentage of smokers drink alcohol?


non_smoke_and_drink <- full_data %>% filter(smoke == 0, drink >= 1) %>% summarise(count = n())
(non_smoke_and_drink$count / non_smoking$count) * 100 # # What percentage of non-smokers drink alcohol?

```

### chisq.test 

```{r}
full_data2 <- full_data

full_data2$smoke <- factor(full_data2$smoke, 
                          levels = c(0, 1, 2), 
                          labels = c("Never", "Before_Smoker", "Still_Smoker"))


full_data2$drink <- factor(full_data2$drink, 
                          levels = c(0, 1, 2), 
                          labels = c("Never", "Rarely", "Frequently"))

table(full_data2$smoke, full_data2$drink)


chisq.test(table(full_data2$smoke, full_data2$drink))


```

Since this p-value is much smaller than 0.05, the result is statistically significant. This indicates that there is a non-random association between smoking and alcohol use.

### visual analyze

```{r}

smoke_drink_df <- as.data.frame(table(full_data2$smoke, full_data2$drink)
)
colnames(smoke_drink_df) <- c("Smoke", "Drink", "Count")


ggplot(smoke_drink_df, aes(x = Smoke, y = Count, fill = Drink)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "smoke and alcohol dist",
         x = "smokeı",
         y = "num of people") +
    theme_minimal()
```

## smoke and alcohol effect on dlact score

### regression model

```{r}
interaction_model <- lm(dlact_score ~ smoke * drink, data = full_data)
summary(interaction_model)

```

drinkFrequently: p = 0.00288 Frequent drinkers have significantly lower DLact scores. Coefficient: -0.37981. This indicates that frequent alcohol consumption reduces DLact scores by approximately 0.38 units.

### visual

```{r}
full_data %>%
  filter(!is.na(smoke), !is.na(dlact_score), !is.na(drink)) %>%
  ggplot(aes(x = factor(smoke), y = dlact_score, color = factor(drink))) +
    geom_boxplot(outlier.shape = NA) +                                    
    geom_jitter(size = 0.2, alpha = 0.2) +                                  
    theme_minimal() +                                                       
    theme(
      legend.position = "right",                                           
      plot.title = element_text(size = 12, face = "bold")
    ) +
    ggtitle("Smoking vs. DLact Score by Alcohol Consumption") +             
    xlab("Smoking Status (0 = Non-smoker, 1 = Smoker)") +
    ylab("DLact Score")

```

### anova and tukey test

```{r}
aov(dlact_score ~ smoke * drink, data = full_data2)
summary(aov(dlact_score ~ smoke * drink, data = full_data2))

TukeyHSD(aov(dlact_score ~ smoke * drink, data = full_data2))

```

Drink appears to have a significant impact on dlact_score, while smoke and the smoke:drink interaction do not seem to show significant effects.

## dlact_score vs mood

```{r}
full_data$mood1 <- factor(full_data$mood1, levels = c("0","1"))


chisq.test(table(full_data$mood1,full_data$dlact_score))
```

A p-value of < 2.2e-16 indicates that the p-value is much smaller than 0.05, meaning that the results are statistically significant.


```{r}
mosaicplot(table(full_data$mood1, full_data$dlact_score))

```

```{r}
full_data %>%
  filter(!is.na(mood1), !is.na(dlact_score)) %>%  
  ggplot(aes(x = factor(mood1), y = dlact_score, fill = factor(mood1))) +
    geom_boxplot() +        
    geom_jitter(color = "black", size = 0.5, alpha = 0.6) +  
    scale_fill_manual(values = c("#FF9999", "#9999FF")) +    
    theme_minimal() +                                       
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      legend.position = "none"
    ) +
    ggtitle("DLact Score by Mood (0 and 1)") +
    xlab("Mood (0 = Negative, 1 = Positive)") +
    ylab("DLact Score")

```


```{r}
anova_result <- aov(dlact_score ~ mood1, data = full_data)
summary(anova_result)
```
## Education Level and Depression

```{r}
table(full_data$isced97total, full_data$mood1)

full_data$mood1 <- as.numeric(full_data$mood1)


anova_result <- aov(mood1 ~ isced97total, data = full_data)
anova_result

```

```{r}
ggplot(full_data, aes(x = isced97total, fill = factor(mood1))) +
  geom_bar(position = "stack") +
  labs(title = " Education Level and Depression",
       x = "ISCED 97",
       y = "frequence",
       fill = "mood (0 = negative, 1 = positive)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## The Relationship Between Working Status and Mood (Emotional State)

```{r}
full_data %>%
  filter(!is.na(p1curocc)) %>%
  ggplot(aes(x = p1curocc, fill = factor(mood1))) +
  geom_bar() +
  labs(title = "The Relationship Between Working Status and Mood", 
       x = "occ..", 
       y = "Frequence",
       fill = "Mood1") +
  theme_minimal()
```


